Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-17 ind_files/01_Introduction.tex
Minor phrasing changes

More than 150 years have passed since Charles Darwin first published \textit{Origin} (in 1859) and Ernst Haeckel first coined the term ‘ecology’ (in 1866). Today, both ecology and evolution are incredibly interdisciplinary fields, borrowing techniques and ideas from diverse areas such as computer science, statistics, economics, dynamical systems, physics, and information theory. With this development has come a cornucopia of models that try to understand biological phenomena in the language of these borrowed tools and techniques. Though many of these ideas are used to accurately describe and analyze specific systems, there is also value to formulating general models in abstract terms that only incorporate a small number of `fundamental' processes and try to capture the `essence' of a biological pattern~\citep{frank_natural_2012, vellend_theory_2016,luque_mirror_2021}. Such general organizing models are vital to theory-building~\citep{luque_mirror_2021} because they help clarify conceptual similarities and unifying factors between apparently disparate modelling frameworks, helping us seamlessly translate essential ideas from one theoretical language to another.

\section{Idealization and generality}\label{idealization}
The push and pull between the search for general patterns and the specification of minutiae has a long and torturous history in ecology~\citep{kingsland_modeling_1985} and evolution~\citep{provine_origins_2001}. The fact that idealization is part and parcel of science is clear if one looks at the actual practice, be it theorists making unrealistic assumptions on paper to model specific phenomena or experimentalists creating artificially controlled conditions in the laboratory to test specific hypotheses~\citep{zuk_models_2018}. Indeed, some philosophers of science recently argued that ``the epistemic goal of science is not truth, but understanding''~\citep{potochnik_idealization_2018}, an idea generally echoed by practicing scientists~\citep{levins_strategy_1966,zuk_models_2018,grainger_empiricists_2022}.  In other words, since the world is complicated and humans are limited, general understanding inevitably comes at the cost of other desirable qualities such as the ability to make precise quantitative predictions. This is especially true for complex phenomena such as those that are the domain of ecology and evolution, where we are often not even aware of all the factors that are at play or how they interact. It is important to remember at the outset that the models I speak about in this thesis may seem to be too abstract/general and may seem to ignore many relevant biological details. This is \textit{by design}, in pursuit of general insight over precise quantitative prediction.

An approach like this will often not be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”\citep{kingsland_modeling_1985}. However, formulating and studying such generalizations can be greatly beneficial as an aid to thinking, sometimes by the very virtue of the fact that it can be done through `blurry eyed' thinking that only looks for general broad-brush regularities\footnote{It bears noting that when a biologist says something is `general', they typically mean that it occurs in the majority of the relevant cases studied. In contrast, when a mathematician says something is `general', they mean that it is \emph{always} true. Thus, an event that occurs in 70\% of cases is `generally true' to biologists but `not true in general' to mathematicians. In this thesis I write as a biologist.}. This is perhaps best illustrated by the success of statistical mechanics in physics - Statistical mechanics was essentially born from the idea that statistical regularities of systems can emerge without the need for knowing the excruciating details of every single moving part, and indeed, starting from first principles, this sort of explicitly `blurry-eyed' thinking was shown to be able to recover the phenomenological laws of thermodynamics as \emph{statistical laws}.

In biology, arguably the greatest such general organizing framework was the idea of evolution by natural selection as synthesized from myriad detailed observations by Charles Darwin. Theoretical population genetics has also had a long-standing tradition in building general organizing frameworks that `abstract away' some biological specificities in favor of a small number of `fundamental' forces, and here, the best analog of statistical mechanics is the Price equation, which was able to recover several known formalisms of evolution from first principles (see section \ref{sec_history} below). In a similar spirit, Vellend has recently argued that conceptual synthesis in community ecology requires
``shifting the emphasis away from an organizational structure based on the useful lines of
inquiry carved out by researchers, to one based on the fundamental processes that underlie community dynamics and patterns''~\citep{vellend_theory_2016}. Vellend's assertion is based on the
fact that population genetics has managed to come up  with reasonably comprehensive theory due to its focus on the abstract `high-level’ processes of selection, mutation, drift, and
gene flow (see section below) instead of the myriad `low-level' processes that may be responsible for generating them. In contrast, he believes that practitioners of community ecology often focus on
specific `low-level' processes such as predation rate, limiting resources (R\textsuperscript{*}), storage effects,
priority effects, senescence, and niche partitioning, leading to a plethora of models (see Table
5.1 in~\cite{vellend_theory_2016} for an in-exhaustive list of 24 such models) and the conclusion that
community ecology `is a mess'. Vellend instead proposes organizing ecological models according to
 the `high-level’ processes of selection, ecological drift (demographic stochasticity), dispersal,
and speciation.

@@

Formal models in classical population genetics often regard forces like selection and mutation as fundamental, and the success of this approach during the Modern Synthesis illustrates the value of formulating high-level, abstract models that only provide a `high-level' description of the fundamental processes required to capture the essence of a biological pattern. However, this evolutionary play famously unfolds in the ecological theatre~\citep{hutchinson_ecological_1965}. Thus, quantities like fitness are not truly fundamental but instead emerge as the net result of various ecological interactions, tradeoffs, and constraints~\citep{metz_how_1992}, a fact that can have important consequences for evolution~\citep{coulson_putting_2006, kokko_can_2017}. Trying to understand such `eco-evolutionary dynamics' has sprouted a rich body of theoretical literature under the broad heading of `evolutionary ecology' and led to the development of theoretical frameworks like evolutionary game theory and adaptive dynamics, fields which have greatly enriched our understanding of biological populations~\citep{brown_why_2016}. Eco-evolutionary population dynamics can (very) broadly be organized under a single unifying framework, the Price equation, that yields all the relevant formal structures (such as evolutionary game theory and classic population genetics) as special cases~\citep{page_unifying_2002, queller_fundamental_2017, lion_theoretical_2018}. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of the high-level evolutionary forces of selection, mutation, and drift, thus providing a useful conceptual framework for thinking about how populations change over time~\citep{frank_natural_2012}. The Price equation also leads to a small number of simple yet insightful `fundamental theorems' of population biology~\citep{queller_fundamental_2017, lion_theoretical_2018, lehtonen_price_2018} and unifies several various seemingly disjoint formal structures under a single theoretical banner~\citep{ lehtonen_price_2020, luque_mirror_2021}

@@

Incorporating stochasticity into deterministic systems is a tricky business, and, if done in a phenomenological manner by adding noise to a `deterministic skeleton'~\citep{coulson_skeletons_2004} in an ad-hoc fashion, can lead to nonsensical predictions and inconsistencies~\citep{strang_how_2019}. Stochastic individual-based models, in which (probabilistic) rules are specified at the level of the individual and population level dynamics are systematically derived from first principles, are self-consistent~\citep{strang_how_2019}, much more natural~\citep{black_stochastic_2012}, and can fundamentally differ from the predictions made by simply adding noise terms to a deterministic model~\citep{strang_how_2019}. Formulating the fundamental formal structures of evolutionary biology in terms of the mechanistic demographic processes of birth and death at the individual level is also greatly desirable for biological reasons~\citep{metcalf_why_2007,geritz_mathematical_2012} because `all paths to fitness lead through demography'~\citep{metcalf_all_2007}. In other words, since demographic processes such as birth and death rates explicitly account for the ecology of the system, they can more accurately reflect the complex interplay between ecological and evolutionary processes and provide a more fundamental mechanistic description of the relevant evolutionary forces and population dynamics~\citep{doebeli_towards_2017}. In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. The mathematical formalism itself is very general and applies equally to the `high level' forces of population genetics as postulated during the Modern Synthesis and the `high level' forces of community ecology as postulated by~\cite{vellend_theory_2016}, though I will mostly stick to the population genetics interpretation in my discussions.

@@

This thesis develops a mathematical formalism for describing finite fluctuating populations from first principles and is structured as follows. The rest of this thesis is divided into two parts. Part \ref{part_theory} provides all the gory mathematical details in a (hopefully) accessible pedagogical style. Part \ref{part_summary} then (re)presents some major results from part \ref{part_theory} and discusses their implications and connections with previous studies.

\begin{itemize}
		\item Chapter \ref{chap_math_background} provides the necessary mathematical background and provides a toy example studying the size of a population of identical individuals that illustrates the major ideas used. If I am required to shoehorn the structure into the `intro-methods-results-discussion' format, then this chapter can be thought of as providing a mathematical `introduction' and illustration of the `methods' that will be used/generalized to get `results' in the following chapters.
		\item Chapter \ref{chap_BD} develops a formalism for describing the evolution of finite fluctuating populations of individuals that come in arbitrarily many `types' which vary in a \emph{discrete} character. This yields equations that generalize the classic Price equation and replicator-mutator equation to finite fluctuating populations.
		\item Chapter \ref{chap_infD_processes} extends the ideas developed in chapter \ref{chap_BD} to populations that vary in a single one-dimensional \emph{quantitative} character and derives a so-called `stochastic field theory' that describes evolution in such populations. This also results in some mathematical equations that may be of independent interest to physicists and applied mathematicians.
		\item Chapter \ref{chap_unification} provides a technical summary of the major results by deriving three important stochastic differential equations. These are equations for type frequencies, the population mean value of an arbitrary type-level quantity, and the population variance of an arbitrary type-level quantity, and respectively generalize the replicator-mutator equation, the Price equation, and \cite{lion_theoretical_2018}'s variance equation to finite, stochastically fluctuating populations. These equations predict a directional evolutionary force called `noise-induced selection' that is only seen in finite, fluctuating populations. Some implications for social evolution and community ecology are discussed. I also briefly discuss the field equation formalism I develop in chapter \ref{chap_infD_processes}.
		\item Chapter \ref{sec_disc} provides a quick summary of the major results and discusses biological implications, connections with previous studies, and opportunities for future work. This chapter has no equations (!). Thus, readers who do not want any mathematical equations can skip all other chapters and directly read chapter \ref{sec_disc} if they are interested only in the final results and takeaways\footnote{Though I promise I have tried my best to make sure the journey to this destination is also pleasant :)} 
\end{itemize}

@@

While the mathematics of chapter \ref{chap_BD} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see~\cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach as applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapters \ref{chap_math_background} and \ref{chap_BD} together also serve as a tutorial and technical introduction to some theoretical ideas: For ecologists, the chapter introduces `system size expansions' and illustrates their use in a general setting, and can be seen as a tutorial on modelling finite populations analytically with minimal assumptions; For population geneticists, the chapter illustrates how system-size approximations (`diffusion approximations' in the population genetics literature) can be carried out without assuming a constant (effective) population size and how this generalization has important consequences for the evolutionary forces at play; For physicists and applied mathematicians, the chapter presents a study of the consequences of applying the system-size expansion to the kind of density-dependent birth-death processes that are widely applicable in ecology and evolution - Unlike many physical systems, though demographic processes like birth and death over ecological timescales are usually formulated in terms of population numbers or densities, predictions in evolution are typically in terms of frequencies of types, and this fact has subtle consequences that can be overlooked if one only works with densities.

Chapter \ref{chap_infD_processes} introduces a function-valued process\footnote{This is really a measure-valued process but we will pretend we don't know this in the interest of accessibility} to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via an analog of the system-size approximation to arrive at a `functional' Fokker-Planck equation in which derivatives are replaced by functional derivatives. I show that classic equations from quantitative genetics such as Kimura's infinite alleles model and Lande's gradient dynamics can be derived as the infinite population limit of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation that can be analyzed for specific systems as required. Unlike the systems studied in Chapter \ref{chap_BD}, formalizing the study of the kind of processes we study in Chapter \ref{chap_infD_processes} is an active area of mathematical research~\citep{carmona_stochastic_1999,da_prato_stochastic_2014,prevot_concise_2007,liu_stochastic_2015,bogachev_fokker-planck-kolmogorov_2015,balan_gentle_2018} and the mathematics itself is far from settled.

Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues~\citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015} to a wide class of eco-evolutionary systems, and to the best of my knowledge, has never been presented in full generality before. Mathematically, chapter \ref{chap_infD_processes} presents heuristic, accessible alternatives to the rigorous tools of martingale theory and measure-valued branching processes that are usually employed to describe the evolution of quantitative traits~\citep{champagnat_unifying_2006,etheridge_mathematical_2011, week_white_2021} by generalizing the idea of a system size expansion of density-dependent (finite-dimensional) birth-death processes to the infinite-dimensional case using the notion of functional differentiation. Biologically, Chapter \ref{chap_infD_processes} provides `stochastic field equations' that describe the dynamics of one-dimensional quantitative traits in finite populations and illustrates that these equations are consistent with well-known formalisms in quantitative genetics at the infinite population limit. Some concrete models are presented as examples in Appendix \ref{App_examples} for clarity regarding the major ideas.

@@

%\section{Coexistence of eco-variants in diverse contexts}\label{synthesis}
%The phenomenon of coexistence of multiple phenotypically distinct, ecologically relevant variants of an organism in sympatry occurs in several different contexts in the natural world. This is best illustrated through examples, which I provide below.
%\subsection{Alternative Reproductive Tactics}
%Alternative reproductive tactics (ARTs) are discrete polymorphisms that occur in sympatry in one or both sexes of a species in response to intrasexual competition for mates. More precisely, two or more traits which are expressed in individuals of a given sex of a given species can be said to be ARTs if~\citep{oliveira_alternative_2008}:
%\begin{enumerate}[label=(\alph*)]
%    \item They are \emph{alternative}, in the sense that a given individual can only express one of the behaviors at any given point in time.
%    \item They are \emph{reproductive}, in the sense that the traits are directly relevant to the process of obtaining a mate and are intended to mitigate conspecific intrasexual competition.
%    \item They are \emph{tactics}, in the sense of serving a well-defined adaptive function and thus having fitness consequences. Usually, different ARTs have different associated fitness effects \emph{ceteris paribus}.
%\end{enumerate}
%ARTs are widespread in the animal kingdom, and may be seen either in morphological traits, or as more complex behavioral polymorphisms (see~\citep{oliveira_alternative_2008} for an overview). Most (if not all) species which exhibit ARTs only show a small number of polymorphisms (usually 2-3, 5 in a handful of cases, and very rarely 7-8) - \hl{I can maybe make a table for this: Somehow wasn't able to find one in the literature}. ARTs have obvious consequences for reproductive life history, and can strongly influence the ecology and evolutionary trajectory of organisms.
%
%\subsection{Trophic resource polymorphisms}
%
%Trophic resource polymorphism is defined as ``the occurrence of discrete intraspecific morphs showing differential niche use, usually through differences in feeding biology and habitat use"~\citep{skulason_resource_1995}. Species which exhibit trophic resource polymorphisms often show marked, discontinuous morphological variations adapted for specific niche usage. For example, arctic charr and african cichlid fishes both often exhibit discrete trophic polymorphisms in which different morphs (of the same species) are specialized for feeding at different strata of the lakes they live in~\citep{recknagel_ecosystem_2017}. Such polymorphisms also generally occur in relatively small numbers in sympatry (see Table 1 in~\citep{smith_evolutionary_1996}). They are relatively widespread in vertebrates, and are thought to have important evolutionary consequences by acting as the `substrate' for adaptive radiations~\citep{smith_evolutionary_1996}. 
%
%\subsection{Mating types in isogamous species}
%Isogamous organisms are sexual, but do not have gametes which can be classified into `male' and `female' based on size. Gametes of such species can, however, be divided into distinct `mating types' according to chemical self-incompatibility. There is substantial variation in the number of mating types present in a given species: While most organisms have only two mating types, this is by no means the rule~\citep{phadke_rapid_2009,constable_rate_2018}, with a handful of organisms having 7-10 mating types, and one species of fungus (\textit{Schizophyllum commune}) exhibiting over 23,000 mating types. The study of mating types in isogamous species is ecologically and evolutionarily relevant, not only due to the intrinsic importance of mating and self-incompatibility, but due to the fact that isogamy is thought to be the evolutionary precursor to anisogamy: Thus, studying the factors governing the number of mating types in isogamous species may shed light on why anisogamous species have only two distinct types of gametes.
%
%\subsection{Phenotypic clustering and sympatric speciation}
%\hl{Write stuff about sympatric speciation, adaptive radiations, and phenotypic clustering.}
%\\
%\\
%In all these cases, we are interested in monitoring the changes in the number of distinct phenotypic entities, in a scenario in which the phenotype is ecologically relevant for fitness. Existing literature has tended to use the word `morph' for this concept, and I will bow down to convention and do the same. However, it is important to remember that in this context, `morphs' are only defined in terms of traits that \textit{affect fitness}, in contrast to the more general use of the word in biology to mean any difference whatsoever (as in single nucleotide polymorphisms, for example). Note that in many of these cases, even though the emergent alternative morphs are discretely distributed, the underlying traits need not be, and indeed are often continuously varying, and only categorized into discrete bins by human experimentalists. Such discrete morphs are often persistent in phylogenetically similar lineages~\citep{jamie_persistence_2020}, suggesting they are evolutionarily predictable to some degree, and are also thought to be important for very general evolutionary processes~\citep{west-eberhard_alternative_1986}. Intraspecific variation also provides an important metric for the diversity harboured by populations and can have important ecological consequences. Two important questions immediately arise:
%
%\begin{enumerate}
%    \item How does a population that is initially monomorphic for a (selectively non-neutral) trait evolve to become polymorphic for this trait in a seemingly discrete manner?
%    \item What governs the number of discrete morphs that co-occur in a population?
%\end{enumerate}
%
%This very general setting motivates a view in which an `individual' is simply characterized by its phenotype as a point in some abstract `trait space'. These questions then concern the distribution of points in this space (analogous to the frequency of each `type' of individual) changes over evolutionary time due to ecological fitness determinants (through the traits that the individuals possess). More formally, we can visualize an abstract trait space $\mathcal{T}$ which captures all possible phenotypes that an individual may exhibit. Any phenotype can then be viewed as some point $x \in \mathcal{T}$. We are interested in the distribution of trait values in the population subject to some ecological and evolutionary rules, and a `morph' or `eco-variant' is just a cluster within this trait space (a mode in the distribution).
% \subsection{Sympatric speciation}
% The origin of species is relatively well understood when populations diverge in allopatry (separated by geographic barriers, for example). In contrast, sympatric speciation, first formally proposed by Maynard Smith \hl{cite}, is much less well-understood, despite empirical evidence that it has occurred in nature \hl{cite}. Since speciation is one of the fundamental processes in

% \section{Frequency-dependent selection}\label{FDS}
% One well-known mechanism of coexistence is `niche partitioning', whereby coexisting types occupy different areas in niche space and thereby reduce competition between types. This of course immediately immediately begs the question of how niche partitioning occurs in the first place. In general, coexistence is also possible without niche partitioning. Such coexistence in the presence of selective pressure (\textit{i.e} when eco-variants are not selectively neutral) occurs through frequency-dependent selection.

>>>>

%More than 150 years have passed since Charles Darwin first published \textit{Origin} (in 1859) and Ernst Haeckel first coined the term ‘ecology’ (in 1866). Today, both ecology and evolution are incredibly interdisciplinary fields, borrowing techniques and ideas from diverse areas such as computer science, statistics, economics, dynamical systems, physics, and information theory. With this development has come a cornucopia of models that try to understand biological phenomena in the language of these borrowed tools and techniques. Though many of these ideas are used to accurately describe and analyze specific systems, there is also value to formulating general models in abstract terms that only incorporate a small number of `fundamental' processes and try to capture the `essence' of a biological pattern~\citep{frank_natural_2012, vellend_theory_2016,luque_mirror_2021}. Such general organizing models are vital to theory-building~\citep{luque_mirror_2021} because they help clarify conceptual similarities and unifying factors between apparently disparate modelling frameworks, helping us seamlessly translate essential ideas from one theoretical language to another.
%
%\section{Idealization and generality}\label{idealization}
The push and pull between the search for general patterns through idealized mathematical models and skepticism about whether these models truly capture important biological details has a long and torturous history in ecology~\citep{kingsland_modeling_1985} and evolution~\citep{provine_origins_2001}. The fact that idealization is part and parcel of science is clear if one looks at the actual practice, be it theorists making unrealistic assumptions on paper to model specific phenomena or experimentalists creating artificially controlled conditions in the laboratory to test specific hypotheses~\citep{zuk_models_2018}. Indeed, some philosophers of science recently argued that ``the epistemic goal of science is not truth, but understanding''~\citep{potochnik_idealization_2018}, an idea generally echoed by practicing scientists~\citep{levins_strategy_1966,zuk_models_2018,grainger_empiricists_2022}.  In other words, since the world is complicated and humans are limited, general understanding inevitably comes at the cost of other desirable qualities such as the ability to make precise quantitative predictions. This is especially true for complex phenomena such as those that are the domain of ecology and evolution, where we are often not even aware of all the factors that are at play or how they interact.

An approach like this will often not be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”~\citep{kingsland_modeling_1985}. MacArthur was speaking primarily about biological generalities and special cases, but in a related vein, if our language of choice for expressing our general events is mathematics, making any non-trivial observations in complex fields such as ecology and evolution often requires \emph{mathematical} approximations and idealizations that may exclude or underplay several `low-level' model-specific details in favor of general description at a `higher' level. Despite this seeming drawback, formulating and studying such generalizations\footnote{It bears noting that when a biologist says something is `general', they typically mean that it occurs in the majority of the relevant cases studied. In contrast, when a mathematician says something is `general', they often mean that it is always true. Thus, a statement that is accurate in 90\% of cases is `generally true' to biologists but `not true in general' to mathematicians. In this thesis I write as a biologist.} can be greatly beneficial as an aid to thinking, sometimes precisely \emph{because} `blurry eyed' thinking that begins from fundamental first principles and only looks for general broad-brush regularities is sometimes more insightful than accounting for every little detail or special case. The success of such an approach is perhaps best illustrated by statistical mechanics in physics - Statistical mechanics was essentially born from the idea that various useful statements about systems with many moving parts can be made without the need for knowing the excruciating details of every single moving part, and indeed, starting from first principles, this sort of explicitly `blurry-eyed' thinking was shown to be able to recover the phenomenological laws of thermodynamics as \emph{statistical laws}.

In biology, arguably the greatest such general organizing framework was the idea of evolution by natural selection as synthesized from myriad detailed observations by Charles Darwin. Theoretical population genetics has also had a long-standing tradition in building general organizing frameworks that `abstract away' some biological specificities in favor of a small number of `fundamental' phenomena like selection and mutation; Here, the most general statistical law we have is the Price equation, which is able to recover several known formalisms of evolution from first principles (see section \ref{sec_history} below). In population ecology, Vellend has recently argued that conceptual synthesis requires
``shifting the emphasis away from an organizational structure based on the useful lines of
inquiry carved out by researchers, to one based on the fundamental processes that underlie community dynamics and patterns''~\citep{vellend_theory_2016}. Vellend's assertion is based on the
fact that population genetics has managed to come up  with reasonably comprehensive theory due to its focus on the abstract `high-level’ processes of selection, mutation, drift, and
gene flow instead of the myriad `low-level' processes that may be responsible for generating them. In contrast, he believes that practitioners of community ecology often focus on specific `low-level' processes such as predation rate, limiting resources (R\textsuperscript{*}), storage effects,
priority effects, senescence, and niche partitioning, leading to a plethora of models (see Table
5.1 in~\cite{vellend_theory_2016} for an in-exhaustive list of 24 such models) and the conclusion that
community ecology `is a mess'. Vellend instead proposes speaking about ecological theory in terms of the `high-level’ processes of selection, ecological drift (demographic stochasticity), speciation,
and dispersal, in direct analogy with the `high-level' processes of selection, genetic drift, mutation, and gene flow respectively in population genetics. Without dwelling on the practicality of such an organization, which has been written about in great detail in~\cite{vellend_theory_2016}, I will simply note that such an organization and broad analogy at the very least seems somewhat plausible from a theoretical viewpoint, and is desirable for providing a cohesive theoretical underpinning for speaking about various hypotheses and ideas in community ecology and how they relate to each other. This is especially relevant given the increasing recognition that population ecology and evolutionary dynamics (Here I really mean `population evolution' as captured by population genetics + quantitative genetics\footnote{Though one could perhaps argue that all evolutionary phenomena are captured in this sort of `population evolution' for some sufficiently general definition of the word `population'. In Michael Lynch's words, ``Nothing in evolution makes sense except in the light of population genetics''~\citep{lynch_origins_2007}. I don't know if this is an extreme view, though, which is why these sentences are in a footnote :)}) are perhaps not as separated as we once thought~\citep{coulson_putting_2006,metcalf_why_2007,schoener_newest_2011,kokko_can_2017,lion_theoretical_2018,govaert_eco-evolutionary_2019,svensson_eco-evolutionary_2019}.

@@

Formal models in classical population genetics often regard forces like selection and mutation as fundamental, and the success of this approach during the Modern Synthesis illustrates the value of formulating high-level, abstract models that only provide a `high-level' description of the fundamental processes required to capture the essence of a biological pattern. However, this evolutionary play famously unfolds in the ecological theatre~\citep{hutchinson_ecological_1965}. Thus, quantities like fitness are not truly fundamental but instead emerge as the net result of various ecological interactions, tradeoffs, and constraints~\citep{metz_how_1992}, a fact that can have important consequences for evolution~\citep{coulson_putting_2006, kokko_can_2017}. Trying to understand such `eco-evolutionary feedbacks' or `eco-evolutionary dynamics' has sprouted a rich body of literature under the broad heading of `evolutionary ecology' and led to the development of theoretical frameworks like evolutionary game theory and adaptive dynamics, fields which have greatly enriched our understanding of biological populations~\citep{brown_why_2016,lion_theoretical_2018,govaert_eco-evolutionary_2019,svensson_eco-evolutionary_2019}. Eco-evolutionary population dynamics can (very) broadly be organized under a single unifying framework, the Price equation, that yields all the relevant formal structures (such as evolutionary game theory and classic population genetics) as special cases~\citep{page_unifying_2002, queller_fundamental_2017, lion_theoretical_2018}. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of the high-level evolutionary forces of selection, mutation, and drift, thus providing a useful conceptual framework for thinking about how populations change over time~\citep{frank_natural_2012}. The Price equation also leads to a small number of simple yet insightful `fundamental theorems' of population biology~\citep{queller_fundamental_2017, lion_theoretical_2018, lehtonen_price_2018} and unifies several various seemingly disjoint formal structures under a single theoretical banner~\citep{ lehtonen_price_2020, luque_mirror_2021}.

@@

Incorporating stochasticity into deterministic systems is a tricky business, and, if done in a phenomenological manner by adding noise to a `deterministic skeleton'~\citep{coulson_skeletons_2004} in an ad-hoc fashion, can lead to models that are not self-consistent~\citep{strang_how_2019}. Further, this procedure of adding noise in an ad-hoc manner provides no insight into the mechanistic factors actually responsible for the stochasticity in the first place. Stochastic individual-based models, in which (probabilistic) rules are specified at the level of the individual and population level dynamics are systematically derived from first principles, are self-consistent, much more natural, and can fundamentally differ from the predictions made by simply adding noise terms to a deterministic model~\citep{black_stochastic_2012,strang_how_2019}. Formulating the fundamental formal structures of evolutionary biology in terms of the mechanistic demographic processes of birth and death at the individual level is also greatly desirable for biological reasons~\citep{metcalf_why_2007,geritz_mathematical_2012} because `all paths to fitness lead through demography'~\citep{metcalf_all_2007}. In other words, since demographic processes such as birth and death rates explicitly account for the ecology of the system, they can more accurately reflect the complex interplay between ecological and evolutionary processes and provide a more fundamental mechanistic description of the relevant evolutionary forces and population dynamics~\citep{doebeli_towards_2017}. In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. The mathematical formalism itself is very general and applies equally to the `high level' forces of population genetics as postulated during the Modern Synthesis and the `high level' forces of community ecology as postulated by~\cite{vellend_theory_2016}, though I will mostly stick to the population genetics interpretation in my discussions.

@@

This thesis develops a mathematical formalism for describing finite fluctuating populations from first principles and is structured as follows. The rest of this thesis is divided into two parts. Part \ref{part_theory} provides the complete formalism in all its gory mathematical detail, but in a (hopefully) accessible pedagogical style. Part \ref{part_summary} then (re)presents some major results from part \ref{part_theory} and discusses their implications and connections with previous studies. Appendix \ref{App_examples} presents some concrete examples of models for clarity regarding the major ideas.

\begin{itemize}
		\item Chapter \ref{chap_math_background} provides the necessary mathematical background and provides a toy example studying the size of a population of identical individuals that illustrates the major ideas used. If I am required to shoehorn the structure into the `intro-methods-results-discussion' format, then this chapter can be thought of as providing a mathematical `introduction' and illustration of the `methods' that will be used/generalized to get `results' in the following chapters.
		\item Chapter \ref{chap_BD} develops a formalism for describing the evolution of finite fluctuating populations of individuals that come in arbitrarily many `types' which vary in a \emph{discrete} character. This yields equations that generalize the classic Price equation and replicator-mutator equation to finite fluctuating populations.
		\item Chapter \ref{chap_infD_processes} extends the ideas developed in chapter \ref{chap_BD} to populations that vary in a single one-dimensional \emph{quantitative} character and derives a so-called `stochastic field theory' that describes evolution in such populations. This also results in some mathematical equations that may be of independent interest to physicists and applied mathematicians.
		\item Chapter \ref{chap_unification} provides a technical summary of the major results by deriving three important stochastic differential equations. These are equations for type frequencies, the population mean value of an arbitrary type-level quantity, and the population variance of an arbitrary type-level quantity, and respectively generalize the replicator-mutator equation, the Price equation, and \cite{lion_theoretical_2018}'s variance equation to finite, stochastically fluctuating populations. These equations predict a directional evolutionary force called `noise-induced selection' that is only seen in finite, fluctuating populations. Some implications for social evolution and community ecology are discussed. I also briefly discuss the field equation formalism I develop in chapter \ref{chap_infD_processes}.
		\item Chapter \ref{sec_disc} provides a quick summary of the major results and discusses biological implications, connections with previous studies, and opportunities for future work. This chapter has no equations (!). Thus, readers who do not want any mathematical equations can skip all other chapters and directly read chapter \ref{sec_disc} if they are interested only in the final results and takeaways. 
\end{itemize}

@@

While the mathematics of chapter \ref{chap_BD} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see~\cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach as applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapters \ref{chap_math_background} and \ref{chap_BD} together also serve as a tutorial and technical introduction to some theoretical ideas: For ecologists, the chapter introduces `system size expansions' and illustrates their use in a general setting, and can be seen as a tutorial on modelling finite populations analytically with minimal assumptions (Also see~\cite{czuppon_understanding_2021} for a similar review that works with tools mostly used in the more `mathematical' literature; I work here with tools mostly used in the physics literature\footnote{Though I do not typically use physics language/notation in this thesis because I am personally closer to mathematics than physics.}); For population geneticists, the chapter illustrates how system-size approximations (`diffusion approximations' in the population genetics literature) can be carried out without assuming a constant (effective) population size and how this generalization has important consequences for the evolutionary forces at play; For physicists and applied mathematicians, the chapter presents a study of the consequences of applying the system-size expansion to the kind of density-dependent birth-death processes that are widely applicable in ecology and evolution - Unlike many physical systems, though demographic processes like birth and death over ecological timescales are usually formulated in terms of population numbers or densities, predictions in evolution are typically in terms of frequencies of types, and this fact has subtle consequences that can be overlooked if one only works with densities.

Chapter \ref{chap_infD_processes} introduces a function-valued process\footnote{This is really something called a measure-valued process, but in the interest of making this thesis accessible to a broader audience, I will not use measure-theoretic language or tools henceforth.} to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via an analog of the system-size approximation to arrive at a `functional' Fokker-Planck equation in which derivatives are replaced by functional derivatives. I show that classic equations from quantitative genetics such as Kimura's infinite alleles model and Lande's gradient dynamics can be derived as the infinite population limit of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation that can be analyzed for specific systems as required. Unlike the systems studied in Chapter \ref{chap_BD}, formalizing the study of the kind of processes we study in Chapter \ref{chap_infD_processes} is an active area of mathematical research~\citep{carmona_stochastic_1999,da_prato_stochastic_2014,prevot_concise_2007,liu_stochastic_2015,bogachev_fokker-planck-kolmogorov_2015,balan_gentle_2018} and the mathematics itself is far from settled.

Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues~\citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015} to a wide class of eco-evolutionary systems, and to the best of my knowledge, has never been presented in full generality before. Mathematically, chapter \ref{chap_infD_processes} presents heuristic, accessible alternatives to the rigorous tools of martingale theory and measure-valued branching processes that are usually employed to describe the evolution of quantitative traits~\citep{champagnat_unifying_2006,etheridge_mathematical_2011, week_white_2021} by generalizing the idea of a system size expansion of density-dependent (finite-dimensional) birth-death processes to the infinite-dimensional case using the notion of functional differentiation. Biologically, Chapter \ref{chap_infD_processes} provides `stochastic field equations' that describe the dynamics of one-dimensional quantitative traits in finite populations and illustrates that these equations are consistent with well-known formalisms in quantitative genetics at the infinite population limit.

@@
